---
title: "Getting Started with scanr"
format: html
vignette: >
  %\VignetteIndexEntry{Getting Started with scanr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

The `scanr` package detects change points in univariate time series with
window-based SCAN statistics.

```{r}
library(scanr)
```

Create a simple series with one mean change around observation 100.

```{r}
set.seed(123)
x <- c(rnorm(100, mean = 0), rnorm(100, mean = 2))
```

Run change-point detection with several window sizes.

```{r}
fit <- scan_cpd(
  x,
  window_sizes = c(20L, 30L),
  n_boot = 100L,
  random_state = 123L,
  change_type = "mean",
  n_jobs = 1L
)

fit
fit$change_points
```

You can compare detected change points against known truth when benchmarking
or testing simulated data.

```{r}
cpd_metrics(
  true_cps = 100L,
  estimated_cps = fit$change_points,
  n = length(x),
  tolerance = 20L
)
```

The package also exposes lower-level statistics.

```{r}
ts_cusum(x[70:130])
ts_wasserstein(x[70:130])
one_wasserstein_distance(x[1:80], x[121:200])
ipm_statistic(x[1:80], x[121:200])
```

Visual helpers make it easier to inspect the detection result.

```{r}
vis_change_points(x, fit, true_change_points = 100L)
vis_vote_scree(fit)
vis_window_votes(fit)
```
